---
title: "Project Draft"
output: html_document
date: "2023-04-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

ncaa_full = read_csv('../../data/ncaa/ncaa_tournament_full.csv')
```

## Project Draft

## Introduction

-   Introduction

    -   The opening sentence or sentences should draw the reader into your report by providing high-level motivation for the analysis. 

    -   Describe the question or questions of interest and the motivation for the analysis.

    -   Conclude the introduction with a **thesis statement**.

        **Thesis**: Higher-seed teams (1-8) make the Final Four more frequently than lower-seeded (9-16) teams, due to a combination of proficient offensive and defensive ratings.

## Background

-   Background

    -   Describe the data set, how it was collected, what the variables mean.

    -   Cite the source of your data.

    -   Describe any background information needed to better comprehend the question or questions you are posing and how the data relate to the question.

    -   Describe any unusual factors which may affect interpretation of results.

    -   Describe what you intend to do in the rest of the report.

## Analysis

-   Analysis

    -   Include numerical and graphical summaries of the data. Be sure to introduce and describe your numerical and graphical summaries.  Do not display a series of summaries without text explaining them.

        -   Including:

            -   Bar graph of [average adjusted offensive rating and defensive rating (AdjOE & AdjDE)]{.underline} by **seed**

            -   Bar graph of [average adjusted strength of schedule (AdjSOS)]{.underline} by **seed**

    -   Make sure to include at least one graphical display of the data that best supports your main conclusions for your primary question, and explain how it supports your conclusions.

    -   For each inference you make, briefly describe the method you use and what the results are.

    -   Provide evidence to support each of your claims.

    -   This section should not include substantial interpretation of results. Leave that for the discussion.

### Offensive Efficiency vs. Defensive Efficiency

```{r}
ggplot(ncaa_full, aes(x = AdjDE, y = AdjOE, color = seed_season)) +
  geom_point() +
  geom_hline(yintercept = mean(ncaa_full$AdjOE)) +
  geom_vline(xintercept = mean(ncaa_full$AdjDE))
```

### Average Adjusted Efficiency Margin by Seed

```{r}
ncaa_seed_sum = ncaa_full %>%
  group_by(seed_season) %>%
  summarize(avg_adj_oe = round(mean(AdjOE), 2),
            avg_adj_de = round(mean(AdjDE), 2),
            avg_adj_em = round(mean(AdjEM), 2))

ggplot(ncaa_seed_sum, aes(x = factor(seed_season), y = avg_adj_em)) +
  geom_col() +
  xlab("Seed") +
  ylab("Adjusted Efficiency Margin")
```

### Top Four Teams (since 2012) in Adjusted Offensive and Adjusted Defensive Efficiency

#### Offensive Efficiency

```{r}
top4_aoe = ncaa_full %>%
  mutate(AdjOE = round(AdjOE, 2)) %>%
  slice_max(order_by = AdjOE, n = 4) %>%
  select(Season, TeamName, AdjOE, seed_season)

top4_aoe %>%
  print.data.frame(digits = 5)
```

> The top four highest adjusted offensive efficiencies since 2012 include **2018 Villanova**, **2015 Wisconsin**, **2021** **Gonzaga**, and **2014 Creighton**. The sample features three one seeds, all of which made the National Championship game.

#### Defensive Efficiency

```{r}
top4_ade = ncaa_full %>%
  mutate(AdjDE = round(AdjDE, 2)) %>%
  slice_min(order_by = AdjDE, n = 4) %>%
  select(Season, TeamName, AdjOE, seed_season)

top4_ade %>%
  print.data.frame(digits = 5)
```

> The top four highest adjusted defensive efficiencies since 2012 include **2013 Louisville**, **2015 Kentucky**, **2018** and **2015** **Virginia**. The sample features three one seeds, two of which made it the Final Four--Louisville being the 2013 National Champion.

## Discussion

-   Discussion

    -   Provide broader interpretations of your analysis and describe how to interpret your results with respect to your question of interest.

    -   Discuss any potential short-comings of the analysis.

    -   Discuss potential future directions for additional work

        -   New questions

        -   Different methods to address the same questions

        -   New data you might collect to refine your understanding

    -   Summarize your primary conclusions and the primary evidence that supports these conclusions.

### References

-   References

In the R Markdown file, you may automatically cite other resources by using a caret followed by the reference between square brackets: \^[your reference text] which will place a marker at this location and a footnote at the end of the document.
